Metric Spaces with Expensive Distances

### 2019-04-21 paper/draft_esa.tex
interim
Fits in page limit

\title{Metric Spaces with Expensive Distances}
\titlerunning{Metric Spaces with Expensive Distances}
% uncomment for LIPICS 
\author{Anonymous Authors}{Anon}{}{}{}
%\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
%\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
%\authorrunning{M. Kerber and A. Nigmetov}
\authorrunning{Anonymous Authors}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\ccsdesc[500]{Theory of computation~Random search heuristics}
\ccsdesc[500]{Theory of computation~Random search heuristics}
%% \category{dummy}

@@

\begin{abstract}
In algorithms for finite metric spaces, it is common
to assume that the distance between two points
can be computed in constant time, and complexity
bounds are expressed only in terms of the number
of points of the metric space.
We introduce a different model where we assume
that the computation of a single distance is
an expensive operation and consequently, the goal is
to minimize the number of such distance queries.
This model is motivated by metric spaces
that appear in the context of topological data analysis.

>>>>

\title{Metric Spaces with Expensive Distances}
\titlerunning{Metric Spaces with Expensive Distances}
% uncomment for LIPICS 
\author{Anonymous Authors}{Anonymous}{}{}{}
%\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
%\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
%\authorrunning{M. Kerber and A. Nigmetov}
\authorrunning{Anonymous Authors}

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\ccsdesc[500]{Theory of computation~Random search heuristics}
%% \category{dummy}

@@

\begin{abstract}
In algorithms for finite metric spaces, it is common
to assume that the distance between two points
can be computed in constant time, and complexity
bounds are expressed only in terms of the number
of points of the metric space.
We introduce a different model, where we assume
that the computation of a single distance is
an expensive operation and consequently, the goal is
to minimize the number of such distance queries.
This model is motivated by metric spaces
that appear in the context of topological data analysis.

### 2019-04-21 paper/expensive-distances-main.tex
Add timings
Fits in page limit

The performance of algorithms for these problems depends on the number of points,
the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
It is a common assumption to assume $\complexity$ to be a constant; 
There are good reasons for that: the most common case of a metric space
is $\metricspace=\RR^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of pairwise distances or runs slighly slower, depending on the approximation
quality.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in brute-force column
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, as done in \dionisus, or
it can be computed with the user-given accuracy using the auction algorithm,
as in \hera. Exact computation is very expensive (see \cite{kerber2017geometry}), and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
In order to make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with a typical cardinality
between 100 and 300 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.

@@

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  0  &  9383\\
$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  0  &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

@@

\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

We also tested higher dimensions and show the results
in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
edges for $700$ points, which clearly shows some degrading for higher dimensions.

The plot in \cref{fig:spanner_eps_dependence}
compares the \bgrdy and \textsc{Greedy} algorithms
on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
that the ratio of the number of edges decreases for
smaller values of $\eps$: 
for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
while for $\eps = 1/32$ the ratio is 2.6

@@

The following geometric lemma
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

\begin{lemma}
\label{lem:bound_lemma}
Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
\begin{enumerate}
\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
of $p_j$ to $q$.
\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
the distance of $p_j$ to $q$.
\end{enumerate}
\end{lemma}
\begin{proof}

>>>>

The performance of algorithms for these problems depends on the number of points,
the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
It is a common assumption to assume $\complexity$ to be a constant.
There are good reasons for that: the most common case of a metric space
is $\metricspace=\RR^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of all pairwise distances or runs slighly slower, depending on the approximation
quality.

@@

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in brute-force column
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, as done in \dionisus, or
it can be computed with the user-given accuracy using the auction algorithm,
as in \hera. Exact computation is very expensive (see \cite{kerber2017geometry}), and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
In order to make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.

@@

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
%$\eps = 0.01$    &   4093  &  0  &&  5331  &  0  &&  16852  &  0\\
$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  11781  &  13557\\
$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  10541   &  9383\\
$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  9418    &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

@@

\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the \bgrdy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

The plot in \cref{fig:blind_greedy_dimensions}
compares performance of \bgrdy in different dimensions.
We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
edges for $700$ points, which clearly shows some degrading for higher dimensions.

The plot in \cref{fig:spanner_eps_dependence}
compares the \bgrdy and \textsc{Greedy} algorithms
on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
that the ratio of the number of edges decreases for
smaller values of $\eps$: 
for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
while for $\eps = 1/32$ the ratio is 2.6

@@

\begin{lemma}
\label{lem:bound_lemma}
Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
\begin{enumerate}
\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
of $p_j$ to $q$.
\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
the distance of $p_j$ to $q$.
\end{enumerate}
\end{lemma}

\cref{lem:bound_lemma}
can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
further distance computations of points very close to $p_i$ or very far from $p_i$
will be avoided.

\begin{proof}
